---
title: "Fuel consumption - Temporal Series Analysis"
author: "Aixa Mas Galiana"
date: "2023-10-21"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warning = F}
# Load libraries, directory & palettes
library(wesanderson)
proj_palette <- c(wes_palette('GrandBudapest2', 4), wes_palette('GrandBudapest1', 4))
directori <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(directori)
```

# Abstract

(Brief decription about the project)

\newpage

# Introduction

In the increasing mobility context given since the last century and around the world, fuel has become a substantial asset(?) in the consumption basket. The analysis of fuel consumption is fundamental in order to understand the influence of this resource nowadays, specifically so that trends and seasonal components can be applied to formulate predictions related to users' consumption.  (+)

Taking into consideration the Consumer Prices Index (CPI; *Índice de precios al consumo*, *IPC*) as a reliable indicator for goods and services prices, the 6.34% of the consumption basket in Spain is associated to fuels and lubricants for personal vehicles. The *indexed* price fluctuations over the years might provide us some valuable information about the fuel consumption oscillations along the time series. 

The following plot elaborated with the same time span as the main time series of the project provides us with the general Consumer Prices Index and the Fuel for personal vehicles Prices Index, using the last year as Index 100. It may be observed that general prices are related with fuel prices (positive fluctuations for the CPI generally result in positive fluctuations for fuel prices) but in a lower scale (fuel price fluctuations tend to be larger, resulting in oscillations of more than 20% in some cases). Even though many other factors should be considered when performing an extended analysis of fuel consumption, observing the trend in price over the years can provide certain key insights on the consumption of this resource. 

> Take into consideration than the series has been deseasonalised through evaluating the values taking as an index the same value for the last year (...) REWRITE.

```{r, message = FALSE, echo = F}
# FUEL 4 PERSONAL CONSUMPTION & GENERIC CPI DATA, INDEX = LAST YEAR PRICE

CPI <- read_excel("./IPC_fuel_anualvar.xlsx", skip = 7, n_max = 4)

CPI_generic <- rev(t(CPI[1,2:133]))
CPI_generic <- window(ts(CPI_generic, start=2009, freq=12), start=2009)

CPI_fuel <- rev(t(CPI[2, 2:133]))
CPI_fuel <- window(ts(CPI_fuel, start=2009, freq=12), start=2009)

plot(CPI_generic, main="Fuel index prices & CPI\nIndex 100 = Last year data", ylab="Index", lwd = 1.5, ylim = c(-30, 35))
lines(CPI_fuel, col = proj_palette[1])
legend("topright", col = c('black', proj_palette[1]), lwd = 2, legend = c('Consumer Price Index (CPI)', 'Fuel for personal consumption Prices Index'), cex = 0.8)
abline(v=2009:2020, lty=2, col = "lightgrey")
abline(h = 0, col = "lightgrey", lty = 2)
```


++++


\ 

## Project overview and objectives

The main objective of this analysis is to detect fuel consumption patterns and build reliable models that take into consideration tendencies and seasonal components using a deterministic approach. 

(Describe what we intend to do in general terms: decomposition of the data in tren, seasonality, etc; interpretation of the trend and causes, interpretation of seasonal indices, etc)

\ 

## Metadata (information about the data)

(Information concerning the data, source, measures, ...)

Consumo de Gasolina en España (Miles de toneladas)

**Fuente**: Ministerio de Fomento. http://www.fomento.gob.es Ministerio de Fomento / Información Estadística / Transporte / Indicadores del transporte

http://www.fomento.gob.es/BE/?nivel=2&orden=18000000

\ 

## Project developement (Road-map)

(Description about the steps followed in each part of the project)


\newpage

# Exploratory Data Analysis (EDA)

## Time Series visualisation

```{r}
serie=window(ts(read.table("gasolina.dat"),start=1993,freq=12),start=2009)

plot(serie, main="Fuel consumption in Spain", ylab="thousands of tonnes")
abline(v=1993:2020,lty=3,col=proj_palette[1])
```

## Summary statistics and Key Insights

\newpage

# Time Series Descomposition


**1. Descriptive centered moving average**: MA with order k = freq/2. We detect the trend & seasonality. 

  - Note we cant use these method to make predictions because it uses observations from the future to elaborate them. It is useful though as it allows us to extract trend deleting seasonality. 

  - LAB AST01 3.3. 
  
  - Also apply 4 & 5 (detrended series, seasonality component estimatation, detrended & deseasonalised series)


**2. Simple exponential smoothing filter (SES)**: allows us to make predictions estimating the smoothing parameter alpha (and later on, a first level estimation).

  - To set alpha we use LAB AST02 5.4 (minimizing MSE, mean square error). We can also use 5.3, finding a grid with the optimal values.
  
  - LAB AST02 6 We make predictions (in and out of sample) 
  
  - Note that in these part we have done just an estimation of alpha. In LAB AST03 2.1.2 we **also estimate the initial value of the level** so that we do not take the first observation as the initial level (l0). We can optimize (MINIMIZE) the MSE function with the SES estimating the l0 and the alpha value. 
  

3. Holt exponential smoothing values: allows us to make predictions taking into consideration the slop (b, *pendiente*) and a smoothing parameter (beta).  

  - LAB AST03 2.2 Complete.
  

4. 
  
  
  


\newpage



# Referencies

IPC on fuel consumption:

https://www.ine.es/jaxiT3/Datos.htm?t=23714#!tabs-tabla





