---
title: "Fuel consumption - Temporal Series Analysis"
author: "Aixa Mas Galiana"
date: "2023-10-21"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
# Load libraries
library(wesanderson)
proj_palette <- c(wes_palette('GrandBudapest2', 4), wes_palette('GrandBudapest1', 4))
```

# Abstract

(Brief decription about the project)

\newpage

# Introduction



```{r, echo = FALSE}
# FUEL & GENERIC DATA WITH INDEX = 2021
CPI_generic <- read_excel("./IPC_generic.xlsx", skip = 7, n_max = 1)
CPI_generic <- rev(t(CPI_generic[26:157]))
CPI_generic <- window(ts(CPI_generic, start=2009, freq=12), start=2009)

CPI_fuel <- read_excel("./IPC_fuel.xlsx", skip = 7, n_max = 3)
CPI_gasoil <- rev(t(CPI_fuel[2,26:157]))
CPI_gasoline <- rev(t(CPI_fuel[3,26:157]))
CPI_fuel <- rev(t(CPI_fuel[1,26:157]))
CPI_fuel <- window(ts(CPI_fuel, start=2009, freq=12), start=2009)
CPI_gasoil <- window(ts(CPI_gasoil, start=2009, freq=12), start=2009)
CPI_gasoline <- window(ts(CPI_gasoline, start=2009, freq=12), start=2009)

plot(CPI_generic, main="Fuel index prices & CPI\nIndex 100 = 2021 data", ylab="Index", lwd = 1.5, ylim = c(80, 180))
abline(v=2009:2020,lty=3,col=proj_palette[2])
lines(CPI_fuel, col = proj_palette[1], lwd = 1.5)
lines(CPI_gasoil, col = proj_palette[3])
lines(CPI_gasoline, col = proj_palette[4])
legend("topright", col = c('black', proj_palette[c(1,3,4)]), lwd = 2, legend = c('Consumer Price Index (CPI)', 'Fuel Prices Index', 'Gasoil Prices Index', 'Gasoline Prices Index'), cex = 0.8)

```

```{r}
# FUEL & GENERIC CPI DATA INDEX = LAST YEAR PRICE

CPI <- read_excel("./IPC_proj.xlsx", skip = 7, n_max = 4)

CPI_generic <- rev(t(CPI[1, 26:157]))
CPI_generic <- window(ts(CPI_generic, start=2009, freq=12), start=2009)

CPI_fuel <- rev(t(CPI[2, 26:157]))
CPI_fuel <- window(ts(CPI_fuel, start=2009, freq=12), start=2009)

CPI_gasoil <- rev(t(CPI[3, 26:157]))
CPI_gasoil <- window(ts(CPI_gasoil, start=2009, freq=12), start=2009)

CPI_gasoline <- rev(t(CPI[4, 26:157]))
CPI_gasoline <- window(ts(CPI_gasoline, start=2009, freq=12), start=2009)


plot(CPI_generic, main="Fuel index prices & CPI\nIndex 100 = Past year data", ylab="Index", lwd = 1.5, ylim = c(-50, 50))
abline(v=2009:2020,lty=3,col=proj_palette[2])
lines(CPI_fuel, col = proj_palette[1])
lines(CPI_gasoil, col = proj_palette[3])
lines(CPI_gasoline, col = proj_palette[4])
legend("topright", col = c('black', proj_palette[c(1,3,4)]), lwd = 2, legend = c('Consumer Price Index (CPI)', 'Fuel Prices Index', 'Gasoil Prices Index', 'Gasoline Prices Index'), cex = 0.8)
```


## Project overview and objectives



## Metadata (information about the data)

## Project developement (Road-map)

(Description about the steps followed in each part of the project)


\newpage

# Exploratory Data Analysis (EDA)

## Time Series visualisation

## Summary statistics and Key Insights

\newpage

# Time Series Descomposition





# Data loading and basic descriptive analysis

Consumo de Gasolina en España (Miles de toneladas)

**Fuente**: Ministerio de Fomento. http://www.fomento.gob.es Ministerio de Fomento / Información Estadística / Transporte / Indicadores del transporte

http://www.fomento.gob.es/BE/?nivel=2&orden=18000000


```{r}
directori <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(directori)

serie=window(ts(read.table("gasolina.dat"),start=1993,freq=12),start=2009)
```

```{r}
plot(serie, main="Consumo de gasolina en España", ylab="miles de toneladas")
abline(v=1993:2020,lty=3,col=wes_palette('GrandBudapest2'))
```

\newpage


# Referencies

https://www.ine.es/jaxiT3/Datos.htm?t=22347



